The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
Second iteration of introduction

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al. 1982; Bézard, B. \& de Bergh, 2007; Way et al. 2016; Way \& Del Genio 2020; ).
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} CITE, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}. CITE
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

We organize the paper as follows:
Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.

>>>>

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet \citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2015AsBio..15..119L,2013ApJ...778..154W,2014ApJ...785L..20W}.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}. Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but can be influenced for different internal redox states and volatile compositions \citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}. However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predicitive theoretical concept and how the habitable zone hypothesis can be tested observationally.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

We organize the paper as follows:
Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
In Section~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
We present our results in Section~\ref{sec:results} before interpreting them in Section~\ref{sec:discussion}.
Finally, we summarize our findings in Section~\ref{sec:conclusions}.
